\bf No More, No Less Than Sum of its Parts: Monoids and The Algebra of Graphics, Statistics, and Interaction

### 2024-10-23 paper.tex
Rewrote some sections

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\hypersetup{
  pdftitle={No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction},
  pdfkeywords={Interactive data visualization, category theory, abstract
algebra, the Gestalt Principles, descriptive statistics},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\if0\blind
{
  \title{\bf No More, No Less Than Sum of its Parts: The Algebra of
Graphics, Statistics, and Interaction}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf No More, No Less Than Sum of its Parts: The Algebra of
Graphics, Statistics, and Interaction}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Interactive data visualization has become a staple of modern data
presentation. Yet, despite its growing popularity, there still remains a
lack of a formal framework for turning raw data into summary statistics
that can then be displayed by interactive graphics. This gap may stem
from a subtle yet profound issue: while we would often like treat
statistics, graphics, and interaction as independent entities, they are
in fact deeply connected. This paper aims to shed light on this
interdependence by relating it to two fundamental concepts from category
theory: groups and monoids. Specifically, if we want our graphics to
support interactive features which break the data into multiple parts
and then combine them back together (such as in the case of linked
selection), then we need the statistics in our plots to fulfill certain
algebraic properties. Ultimately, understanding this hidden structure
may empower data visualization experts to build more flexible and
expressive interactive data visualization systems.
\end{abstract}

@@

The essence of data visualization is comparison. When we visualize, we
split our data into parts, summarize each part one or more summary
statistics, encode the summaries into visual attributes such as
position, size, or color, and finally use these attributes to draw
geometric objects
\citetext{\citealp{bertin1983}; \citealp{wilkinson2012}; \citealp[for a
recent review, see][]{franconeri2021}; \citealp{wilke2019}}. For
example, when drawing a typical barplot, we split our data into parts
based on the levels of some categorical variable, count the number of
cases in each part, and finally represent the counts by drawing bars of
the corresponding heights. In this way, each bar becomes a visual
representation of its own ``small data set'', a subset of the original
data.

To ground this idea in a more concrete example that will be used
throughout this paper, imagine going through your bookshelf, sorting the
books by genre, and stacking them on top of each other. You would
construct a physical version of the barplot, forming one ``bar'' (stack)
per genre: one bar for sci-fi novels, one bar for romance, one bar for
detective fiction, and so on. Even from this very simple example,
interesting mathematical properties jump out: for example, a single book
cannot go into two stacks (the subsets of the data are \emph{disjoint})
and if all of the stacks are combined back together, we should end up
with the bookshelf we started with \citep[the subsets are
\emph{surjective} and cover the entirety of the data set,
see][]{ziemkiewicz2009}.

Now, just as we can learn about the contents of our library by
organizing books in the way we just described, we can learn about our
data by splitting it into parts. A natural question then arises: can we
learn more by subdividing our data into even smaller parts and comparing
these? For instance, in our book example, we can split each stack of
books into two smaller stacks based on the gender of the author. Each
smaller stack will then represent the unique combination (Cartesian
product) of the two categorical variables (genre and gender). Moreover,
there arises a hierarchical relationship between the stacks: by
combining two genre-gender stacks (child components) together, we can
recover the original genre stack (parent component). Similar situation
often arises in interactive graphics, where we use techniques such as
linked selection to highlight specific subsets of our data. Either way,
by forming these finer partitions, we can represent our data with
greater granularity and discover trends that may otherwise hide in the
aggregate.

However, once we have subdivided each category into its child
components, we need to represent the information in a way that visually
preserves the hierarchical nature of this relationship. In data
visualization, there are two popular methods for doing this, known as
``stacking'' and ``dodging'' (see Figure \ref{fig:playfair}). These
methods have existed at least since the time of William Playfair
\citetext{\citeyear{playfair1801}; \citeyear{playfair1822}}. Let's
briefly illustrate them on the example of the classical barplot. In a
stacked barplot, bar segments representing the product of the two factor
variables are plotted vertically on top of each other (``stacked''; the
operation can be also thought of as ``highlighting'' parts of the bar).
In contrast, in a dodged barplot, bars are plotted side-by-side, as
clusters of bars. Finally, there is also a third popular method, known
as layering, where we plot bar segments in separate graphical layers and
use partial transparency to mitigate overplotting (this technique may
have been a bit challenging for William Playfair but has been made
simple with computers).

@@

\caption{Examples of the three methods for dealing with bars that have subdivided into segments. Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822). Bottom row: layering (created with ggplot2; Wickham, 2010).}\label{fig:playfair}
\end{figure}

Much has been written about the relative merits of stacking, dodging,
and layering. For example, layering is only useful with few categories,
as blending many colors can make it difficult to tell the categories
apart \citep{franconeri2021, wilke2019}. Further, in a landmark study,
\citet{cleveland1984} showed that people tend to be less accurate when
reading information from stacked bar charts as opposed to dodged bar
charts. Specifically, since the lower y-axis coordinate of a stacked
segment is pushed up by the cumulative height of the segments below it,
it becomes difficult to accurately compare segments' length, both within
and across bars \citep{cleveland1984}. Subsequent research has
independently validated these findings and expanded upon them \citep[see
e.g.][]{heer2010, thudt2016, quadri2021}. Ultimately, the suboptimal
statistical legibility of stacking has led some authors to advise
against its use altogether \citep{kosara2016, wilke2019}, or at least
urge caution \citep[see e.g.][]{byron2008, cairo2014, franconeri2021}.

@@

Over the course of our evolutionary history, the human brain has evolved
a sophisticated pipeline for processing visual information. This
pipeline has been the subject of a great body of research in cognitive
neuroscience
\citetext{\citealp{goebel2004}; \citealp{knudsen2020}; \citealp[for a
brief overview, see][20]{ware2019}}. Many of the findings from this
research are directly applicable to data visualization. For example,
certain particularly salient visual stimuli are perceived automatically,
rapidly (\textless10 ms), in parallel, and at a constant speed
regardless of the number of distractors \citep{treisman1985}. This
phenomenon, known as pre-attentive processing, is key for data
visualization, as the right use of visual attributes such as color,
contrast, or shape can help us design figures which communicate insights
much more effectively \citep{ware2019}.

However, many of the findings from visual perception research are quite
specific; to guide effective visualization, we also need some general,
high-level rules. One such set of high-level rules are the Gestalt
principles of visual perception \citep[see
e.g.][]{cairo2012, ware2019, rosli2015, vanderplas2020, todorovic2008, pinker1990}.
These principles formulate a set of fundamental ``laws'' related to the
perception of patterns, primarily visual. They were discovered by the
founders of the Gestalt school of psychology early in the 20th century,
primarily by means of self-observation, and, despite not being grounded
in sound understanding of the neural mechanisms of the brain, have
nevertheless been later independently supported by findings from
cognitive neuroscience \citep{guberman2017, ware2019, wertheimer1938}.

Among the Gestalt principles, there are a few which are particularly
relevant to data visualization. First, there are the two principles of
proximity and similarity. These principles state that objects that which
appear in close together and/or similar in appearance tend to be
perceived as part of single group. Second, the closure and common region
principles state that regions with closed contours tend to be perceived
as single, unified objects. Finally, the figure and ground principle
states we tend to see one part of a scene as the central object (the
``figure'') while the rest is perceived as the background (the
``ground''). This can be modulated by other Gestalt principle: for
example, smaller, darker, and symmetric objects with closed contours are
more likely to be perceived as the figure \citep{ware2019}.

Whether we intend them to or not, the Gestalt principles affect how our
data visualizations are perceived. As such, it is a good idea to try to
use them to our advantage. Sometimes, little effort is needed. Such is
the case, for example, when identifying clusters in a scatterplot
\citep{rosli2015}: the principles of proximity and similarity ensure
that clouds of similarly shaped and colored points will be perceived as
discrete groups by the observer. In other cases, however, our design
choices can have a significant impact. For instance, another way to have
objects to be perceived as part of a group is to draw a closed shape
around them \citep{cairo2012}. Alternatively, if we want to highlight a
particular part of a data set, we can leverage the figure and ground
principle and use attributes such as color and saturation to make it
stand out \citep[for specific examples,
see][34]{franconeri2021, wilke2019}.

Finally, ignoring or violating the Gestalt principles can make our
graphics less legible or even misleading. \citet{tufte2001} had
documented many such violations, such as deceptive 3D effects,
inadvertent optical illusions caused by dense stripe patterns, and the
overuse of superfluous graphical elements he termed ``chartjunk''.
Similarly, \citet{cairo2019} highlights many cases of graphics that
violate the Gestalt principles. Finally, there is also the issue of
well-designed graphics that communicate no statistical information at
all, or even information that is blatantly false
\citep{cairo2014, franconeri2021, cairo2019, gelman2013, tufte2001}.
Whether intentional or not, poor design can have a significant impact on
the legibility and effectiveness of statistical graphics.

@@

The Gestalt principles apply to both static and interactive
visualizations, however, the way they apply is different. Static data
visualizations can be designed well or poorly, but this design quality
does not change over time. In contrast, interactive visualizations
change over time, with user interaction. As such, an interactive
visualization may conform to the Gestalt principles at one instance of
time and violate them in another. In fact, interactive visualizations
can even differ in how well they leverage design principles throughout
time: the animations that interactions induce may themselves conform to
good or poor design \citep[see][]{hullman2013, ware2019}.

Thus, interactivity inherently places further constraints the design of
figures. If we want well-designed interactive figures, we need to ensure
that they conform to good design principles across all states induced by
user-interaction. This may not always be easy. We will investigate this
idea in the following section, on the example of linked selection.

@@

While these constraints may not seem too restrictive, we argue that they
pose significant challenges for certain types of plots and visualization
styles. Specifically, some kinds of plots may produce undesirable visual
behavior when combined with linked selection. In fact, this may explain
why stacking is preferred over dodging within the interactive data
visualization community. Let's explore this idea in more depth by
comparing linked selection with a stacked vs.~dodged barplot.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/display-1.pdf}
\caption{\label{fig:display}The visual coherence of stacking vs.~dodging
during linked selection on the \texttt{mtcars} data set
\citep{henderson1981}. Bars show the numbers of cars in the 4-, 6-, and
8-cylinder categories. Dark segments represent selected cases, light
segments represent unselected cases, and the plots' contour is shown via
a light outline. Plots left to right show more cars being selected. Top
row: the contour of a stacked barplot remains constant throughout
selection (three bars of the same height). Bottom row: the contour of a
dodged barplot changes dramatically with selection. Note the changes to
the upper y-axis limit and the absence of the lighter segment among the
6-cylinder cars in the rightmost plot (since all 6-cylinder cars are
selected, the darker segment has been stretched to fill the available
space; \texttt{ggplot2} default for handling empty subcategories).}
\end{figure}

First, we propose that stacking is the superior method for
\emph{displaying} selection (see Figure \ref{fig:display}). In a stacked
barplot, each stacked bar presents itself as a single visual entity,
thanks to the four Gestalt principles: proximity, similarity, closure,
and common region. Since the stacked segments are placed right on top of
each other, and share the same width and closed border, they are
perceived as part of a unified whole (the bar). This remains true
throughout selection - no matter how the heights of the highlighted
segments within a stacked bar change, the height of the whole bar
remains constant, and so does the overall outline of the plot.

Dodged barplots lack this degree of visual consistency. In a dodged
barplot, bar segments placed side-by-side in clusters, and, since each
segment can be of different length, these clusters lack a shared border
(common region). This leads to several design challenges. First, when
selection occurs, the jagged contour of the dodged barplot can change
dramatically as more cases are added or removed. This creates
distracting visual clutter that makes it difficult to track changes over
time. Second, since the heights of the bar segments are not bound by any
value present in the plot (unlike the heights of the segments in the
stacked barplot, which are bound by the height of the tallest bar), we
need to decide how the handle the upper y-axis limit in response to
selection. One option is to make the upper y-axis limit dynamic. If we
do this, we ensure that the bar segments always fit the plotting region,
however, we also lose the context that the upper y-axis limit provides
each time selection occurs. Alternatively, we can assign a static value
to the upper y-axis limit, however, then we risk the bar segments
growing outside of the bounds of the plotting region.

Stacked barplots offer several other advantages over dodged bars within
the context of linked selection. First, we propose that stacking also
provides a simpler mental model for \emph{selection}. Since each stacked
bar is a single visual entity, it makes intuitive sense that clicking it
will select the entire category corresponding to it. Indeed, this is how
many interactive data visualization systems implement linked selection:
clicking on part of an object selects the category corresponding to the
entire object \citep[see
e.g.][]{cook2023, vegalite2022, vanderplas2018, xie2014, theus2002, urbanek2003, urbanek2011, swayne2003}.
In dodged barplots, on the other hand, the expectation is less clear:
when we click on a bar segment in a dodged barplot, will we select the
entire category or just the cases corresponding to the segment? Second,
in a stacked barplot, the bar width and inter-bar gap are constant
throughout selection. In a dodged barplot, it is not clear how to handle
the gaps when some selection groups are empty. For example, when no
cases are selected, should we leave gaps for the (currently absent)
highlighted segments, or draw wide bars and only shrink these in
response to selection? Finally, by presenting fewer visual entities on
the screen, stacked bars may also reduce cognitive load
\citetext{\citealp{sweller2019}; \citealp[for a non-technical overview,
see][]{knaflic2015}}.

To summarize, stacked bars respond gracefully to selection. In contrast,
dodging produces several undesirable behaviors when combined with
selection: it produces jagged objects with non-uniform outline that
changes dramatically with selection. These difference may explain why
stacking is still popular in the interactive data visualization
community.

@@

While data visualizations are made up of geometric objects such as
points, lines, or areas, these objects alone are not enough to
constitute data visualizations. A coloured rectangle holds no meaning in
and of itself; a canvas filled with such rectangles is just an image
(see Figure \ref{fig:magritte}). To turn an image of rectangles into a
plot, we need to make sure that each rectangle accurately represents
some underlying data.

@@

\caption{An image is not the thing it represents. Left: The Treachery of Images by René Magritte (1929) makes the point that an image of a thing is different from the thing itself. Right: similarly, an image containing a few coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data.}\label{fig:magritte}
\end{figure}

However, we can rarely just ``throw'' data at the computer screen, in
its raw form. Instead, we often first need to translate it into a format
that the graphics device can understand. In most data visualization
packages, this is done in two steps: aggregation and scaling.

First, the data needs to be aggregated or ``wrangled'' into sets of
statistical summaries. These summaries may be very simple: for instance,
in a typical scatterplot, we assign each point the raw values of the x-
and y-variables (the summary function is then just identity,
\(f(x) = x\)). More often, however, the data is grouped and aggregated
in some way. For example, as was discussed previously, in a barplot,
data points are grouped by the levels of the x-axis variable, and each
group is then summarized by some descriptive statistic, such as count,
sum, mean, or maximum. Similarly, in a typical histogram, the x-axis
variable is divided into bins, and each bin is then summarized by the
number of cases that fall inside it.

Second, the computed statistics have to be encoded into graphical
attributes such as position, length, size, or colour \citep[see
e.g.][]{franconeri2021, wilkinson2012, pinker1990}. This is the job of
specialized functions known as scales or coordinate systems \citep[see
e.g.][]{murrell2005, murrell2007, wickham2010, wickham2016, wilkinson2012}.
For instance, in a typical scatterplot, the values of the x-axis
variable are mapped to the x-position, such that points corresponding to
small values of the x-axis variable are placed near the left-hand side
of the plot (and larger values are place near the right-hand side).
These mappings may be subject to non-linear transformations, such as log
or square-root, and may also apply to discrete values such as barplot
categories. Scales and coordinate systems are a rich subject and we have
barely scratched the surface here; interested reader is advised to see
\citet{wilkinson2012}, \citet{ziemkiewicz2009}, or \citet{hotz2020}.

@@

What does stacking mean in the context of the statistics we are trying
to represent? To a novice, stacking may appear as a purely graphical
operation. If they attempt to implement a data visualization pipeline
from scratch, they may leave stacking to the very end, after the
statistical summaries have already been translated into screen
coordinates.

This view of stacking as a purely graphical operation has a certain
appeal. Many modern data visualization systems follow the grammar-based
model of visualization
\citep{mcnutt2022, kim2022, vanderplas2020, wickham2010, satyanarayan2014, satyanarayan2016, wilkinson2012}.
This model emphasizes compositionality, allowing plots to be built from
independent, modular components, which the users can combine freely. For
example, in the popular \texttt{ggplot2} library \citep{wickham2010},
plots are built by combining components such as geometric objects
(called \texttt{geoms}) and statistical summaries (\texttt{stats}).
These grammar-based approach offers a great deal of expressiveness, and
has gained widespread popularity as a result.

Thus, it may be tempting to treat stacking just as another modular
component of a visualization, such that we may be able to stack any
geometric objects and statistics we like. Unfortunately, this approach
is deeply flawed. Many data visualization experts have issued explicit
warnings about this:

@@

The message is clear: stacking some statistics such as counts, sums, and
percentages is acceptable, but stacking others is not. As outlined by
Wills, while the sum of grouped counts produces a valid overall count,
the sum of group means does not yield a meaningful statistic. Even
averaging group means, as noted by \citet{wu2022}, does not produce a
valid summary - the mean of group means differs from the grand mean
(unless by coincidence). This problem is extends beyond barplots; for
example, concerns about the statistical validity of stacked density
plots have also been raised recently \citep{pu2020}.

>>>>

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\hypersetup{
  pdftitle={No More, No Less Than Sum of its Parts: Monoids and The Algebra of Graphics, Statistics, and Interaction},
  pdfkeywords={Interactive data visualization, category theory, abstract
algebra, the Gestalt Principles, descriptive statistics},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\if0\blind
{
  \title{\bf No More, No Less Than Sum of its Parts: Monoids and The
Algebra of Graphics, Statistics, and Interaction}

@@

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf No More, No Less Than Sum of its Parts: Monoids and The
Algebra of Graphics, Statistics, and Interaction}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Interactive data visualization has become a staple of modern data
presentation. Yet, despite its growing popularity, there still remains a
lack of a formal framework for turning raw data into summary statistics
that can then be displayed by interactive graphics. This gap may stem
from a subtle yet profound issue: while we would often like treat
statistics, graphics, and interaction as independent entities, they are
in fact deeply connected. This paper aims to shed light on this
interdependence by relating it to two fundamental concepts from category
theory: groups and monoids. Specifically, if we want our graphics to
support interactive features which split our data into parts and then
combine these parts back together (such as linked selection), then we
need the statistics in our plots to fulfill certain algebraic
properties. Ultimately, understanding this hidden structure may empower
data visualization experts to build more flexible and expressive
interactive data visualization systems.
\end{abstract}

@@

The essence of data visualization is comparison. When we visualize, we
split our data into parts, summarize each part by one or more summary
statistics, encode the summaries into visual attributes such as
position, size, or color, and, finally, use these attributes to draw
geometric objects
\citetext{\citealp{bertin1983}; \citealp{wilkinson2012}; \citealp[for a
recent review, see][]{franconeri2021}; \citealp{wilke2019}}. For
example, when drawing a typical barplot, we split our data into parts
based on the levels of some categorical variable, count the number of
cases in each part, and finally represent these counts by drawing bars
of corresponding heights. In this way, each geometric object (bar)
becomes a representation of its own small data set, a subset of the
original data.

To ground this idea in a concrete example that will be used throughout
this paper, imagine going through your bookshelf, sorting the books by
genre, and stacking them on top of each other. You will construct a
physical version of the barplot, forming one ``bar'' (stack) per genre:
one bar for sci-fi novels, one bar for romance, one bar for detective
fiction, and so on. Even this very simple example reveals some
interesting mathematical properties. For example, a single book cannot
go into two stacks (the subsets of the data are \emph{disjoint}) and if
all of the stacks are combined back together, we should end up with the
bookshelf we started with \citep[the subsets are \emph{surjective} and
cover the entirety of the data set, see][]{ziemkiewicz2009}.

Now, just as we can learn about the contents of our library by sorting
books into stacks, we can learn about our data by splitting it into
parts. A natural question then arises: can we learn more by subdividing
our data into even smaller parts? For instance, in our book example, we
can split each stack of books into smaller stacks based on the gender of
the author. Each smaller stack then represents the unique combination
(Cartesian product) of the two categorical variables (genre and gender).
Moreover, there arises a hierarchical relationship between the stacks:
by combining genre-gender stacks (child components), we can recover the
original genre stack (parent component). Similar situation often arises
in interactive graphics, where we use techniques such as linked
selection to highlight specific subsets of our data. Either way, by
forming these finer partitions of our data, we can represent it with
greater granularity and discover trends that may otherwise hide in the
aggregate.

However, once we have subdivided each category into its child
components, we need to represent the information in a way that visually
preserves this hierarchical relationship. In data visualization, there
are two popular methods for doing this, known as ``stacking'' and
``dodging'' (see Figure \ref{fig:playfair}). These methods have existed
at least since the time of William Playfair
\citetext{\citeyear{playfair1801}; \citeyear{playfair1822}}. Let's
briefly illustrate them on the example of the barplot. In a stacked
barplot, bar segments representing the product of the two factor
variables are plotted vertically on top of each other (``stacked''; the
operation can be also thought of as ``highlighting'' parts of the bar).
In contrast, in a dodged barplot, bars are plotted side-by-side, as
``clusters''. Finally, there is also a third more recent method, known
as layering, where we plot bar segments in separate graphical layers and
use partial transparency to mitigate overplotting.

@@

\caption{Examples of three methods for dealing with bars that have split into child components (segments). Top row: dodging (left) and stacking/highlighting (right); both by William Playfair (1801, 1822). Bottom row: layering (created with ggplot2; Wickham, 2010).}\label{fig:playfair}
\end{figure}

Much has been written about the relative merits of stacking, dodging,
and layering. For example, layering is only useful with few categories,
as blending many colors can make it difficult to tell the categories
apart \citep{franconeri2021, wilke2019}. Further, in a landmark study,
\citet{cleveland1984} showed that people tend to be less accurate when
reading information from stacked bar charts as opposed to dodged bar
charts. Specifically, since the lower y-axis coordinate of a stacked
segment is pushed up by the cumulative height of the segments below, it
becomes difficult to accurately compare segments' length, both within
and across bars \citep{cleveland1984}. Subsequent research has
independently validated these findings and expanded upon them \citep[see
e.g.][]{heer2010, thudt2016, quadri2021}. Due to this suboptimal
statistical legibility, many data visualization researchers have urged
caution about stacking \citep[see
e.g.][]{byron2008, cairo2014, franconeri2021}, and some have even
discouraged its use altogether \citep{kosara2016, wilke2019}.

@@

Over the course of our evolutionary history, the human brain has evolved
a sophisticated network for processing visual input. This visual
processing pipeline has been the subject of a great body of research in
cognitive neuroscience
\citetext{\citealp{goebel2004}; \citealp{knudsen2020}; \citealp[for a
brief overview, see][20]{ware2019}}, and many of its findings are
directly applicable to data visualization. For example, certain
particularly salient visual stimuli are perceived automatically, rapidly
(\textless10 ms), in parallel, and at a constant speed regardless of the
number of distractors \citep{treisman1985}. This phenomenon, known as
pre-attentive processing, is key for data visualization, as the right
use of visual attributes such as color, contrast, or shape can help us
design figures which communicate insights much more effectively
\citep{ware2019}.

However, many of the visual perception findings are quite specific and
niche; to guide effective visualization, we also need some general,
high-level rules. One such set of high-level rules are the Gestalt
principles of visual perception \citep[see
e.g.][]{cairo2012, ware2019, rosli2015, vanderplas2020, todorovic2008, pinker1990}.
These principles formulate a set of fundamental ``laws'' related to the
perception of (primarily visual) patterns. They were discovered by the
founders of the Gestalt school of psychology early in the 20th century,
primarily by means of self-observation, and, despite not being grounded
in sound understanding of the neural mechanisms of the brain, have
nevertheless been later independently supported by findings from
cognitive neuroscience \citep{guberman2017, ware2019, wertheimer1938}.

Among the Gestalt principles, there are a few which are particularly
relevant to data visualization. First, there are the two principles of
proximity and similarity. These principles state that objects that which
appear in close together and/or similar in appearance tend to be
perceived as part of single group. Second, the closure and common region
principles state that regions with closed contours tend to be perceived
as single, unified objects. Finally, the figure and ground principle
states that we tend to perceive one part of a visual scene as the
central object (the ``figure'') and the rest as background (the
``ground''). This can be modulated by other Gestalt principle: for
example, smaller, darker, and symmetric objects with closed contours are
more likely to be perceived as the figure \citep{ware2019}.

Whether we intend them to or not, the Gestalt principles affect how our
data visualizations are perceived. As such, it is a good idea to try to
use them to our advantage. Sometimes, little effort is needed. Such is
the case, for example, when identifying clusters in a scatterplot
\citep{rosli2015}: the principles of proximity and similarity ensure
that clouds of similarly shaped and colored points will be perceived as
discrete groups. In other cases, however, our design choices can have a
significant impact. For instance, another way to make objects be
perceived as part of a group is to draw a closed shape around them
\citep{cairo2012}. Alternatively, if we want to highlight a particular
part of a data set, we can leverage the figure and ground principle and
use attributes such as color and saturation to make it stand out
\citep[for specific examples, see][34]{franconeri2021, wilke2019}.

Finally, ignoring or violating the Gestalt principles can make our
graphics less legible or even misleading. \citet{tufte2001} had
documented many such violations, such as deceptive 3D effects,
inadvertent optical illusions caused by dense stripe patterns, and the
overuse of superfluous graphical elements termed ``chartjunk''.
Similarly, \citet{cairo2019} highlights many cases of graphics that
violate the Gestalt principles. Finally, there is also the issue of
well-designed graphics that communicate no statistical information at
all, or even information that is blatantly false
\citep{cairo2014, franconeri2021, cairo2019, gelman2013, tufte2001}.
Whether intentional or not, poor design can have a significant impact on
the legibility and effectiveness of statistical graphics.

@@

The Gestalt principles apply to both static and interactive
visualizations, however, the way they apply is different. Static data
visualizations can be designed well or poorly, but this design quality
does not change over time. In contrast, interactive visualizations
change over time, with user interaction. As such, an interactive
visualization may conform to the Gestalt principles at one instance of
time and violate them in another. In fact, interactive visualizations
can even differ in how well they leverage design principles
\emph{across} time: the animations that interactions induce may
themselves conform to good or poor design
\citep[see][]{hullman2013, ware2019}.

Thus, interactivity inherently places additional constraints the design
of our figures. If we want well-designed interactive figures, we need to
ensure that they conform to good design principles across all states
induced by user-interaction. This may not always be easy. We will
investigate the design challenged posed by interaction in the following
section, on the example of linked selection.

@@

While these constraints may not seem too restrictive, we argue that they
pose significant challenges for certain visualization styles.
Specifically, some kinds of plots may produce undesirable visual
behavior when combined with linked selection, and this may in fact
explain why stacking is preferred over dodging within the interactive
data visualization community. Let's explore this idea in more depth by
comparing linked selection with a stacked vs.~dodged barplot.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/display-1.pdf}
\caption{\label{fig:display}The visual coherence of stacking vs.~dodging
during linked selection. Dark segments represent selected cases, light
segments represent unselected cases, and the plots' contour is shown via
a light outline. Plots left to right show more cases being selected. Top
row: the contour of a stacked barplot remains constant throughout
selection (three bars of the same height). Bottom row: the contour of a
dodged barplot changes dramatically with selection. Note the changes to
the upper y-axis limit and the absence of the lighter segment among the
bar representing category B in the rightmost plot (the darker segment
has been stretched to fill the available space; \texttt{ggplot2} default
for handling empty subcategories). Data source: \citet{henderson1981}.}
\end{figure}

First, we propose that stacking is the superior method for
\emph{displaying} selection (see Figure \ref{fig:display}). Thanks to
the four Gestalt principles of proximity, similarity, closure, and
common region, in a stacked barplot, each bar presents itself as a
single visual entity. In other words, since the stacked segments are
placed right on top of each other, and share width and a closed border,
they are perceived as part of a unified whole (the bar). This remains
true throughout selection - no matter how the heights of the highlighted
segments change, the height of the whole stacked bar remains constant,
and so does the overall outline of the plot.

Dodged barplots lack this degree of visual consistency. In a dodged
barplot, bar segments placed side-by-side in clusters, and, since each
segment can be of different length, these clusters lack a shared border
(common region). This leads to several design challenges. First, when
selection occurs, the jagged contour of the dodged barplot can change
dramatically as cases are added or removed. This creates distracting
visual clutter that makes it difficult to track changes over time.
Second, since the heights of the bar segments are not bound by any value
present in the plot (unlike the heights of the segments in the stacked
barplot, which are bound by the height of the tallest bar), it is not
clear how to handle the upper y-axis limit. On one hand, we could make
the upper y-axis limit dynamic and have it change in response to
selection. If we do this, we ensure that the bar segments always fit the
plotting region, however, we also lose the context that the limit
provides each time selection occurs. Alternatively, we can set the limit
to a static value, however, then we are at the risk of the bar segments
growing outside of the bounds of the plotting region.

Stacked bars may offer several other advantages over dodged bars within
the context of linked selection. First, we propose that stacking also
provides a simpler mental model for \emph{selection}. Since each stacked
bar is a single visual entity, it makes intuitive sense that clicking it
will select the entire category corresponding to it. Indeed, this is how
many interactive data visualization systems implement linked selection:
clicking on part of an object selects the category corresponding to the
entire object \citep[see
e.g.][]{cook2023, vegalite2022, vanderplas2018, xie2014, theus2002, urbanek2003, urbanek2011, swayne2003}.
On the other hand, with dodged bars, the expectation is less clear: when
we click on a segment in a dodged barplot, do we select the entire
category or just the cases corresponding to the segment? Second, in a
stacked barplot, the bar width and inter-bar gap are constant throughout
selection. In a dodged barplot, it is not clear how to handle the gaps
when some selection groups are empty - should we leave gaps for the
(currently absent) highlighted segments, or draw wide bars and only
shrink them in response to selection? If we do the former, the barplot
may show asymmetric gaps and illusory clustering due to the absence of
some segments. If we do the latter, we break the principle that the area
of the bar should be proportional to the number of cases it represents
\citep{cleveland1985}. Finally, by presenting fewer visual entities on
the screen, stacked bars may also reduce cognitive load
\citetext{\citealp{sweller2019}; \citealp[for a non-technical overview,
see][]{knaflic2015}}.

To summarize, stacked bars exhibit a more graceful and visually pleasing
behavior when combined with selection. In contrast, dodging produces
non-uniform objects with jagged outline that changes significantly with
selection. These difference may explain why stacking is still popular in
the interactive data visualization community.

@@

Data visualizations are made up of geometric objects such as points,
lines, or areas. However, these objects alone hold no meaning. Without
context, an image filled with coloured rectangles is just that - an
image (see Figure \ref{fig:magritte}). To turn the into a plot, we need
to make sure that each rectangle accurately represents some underlying
data.

@@

\caption{A representation is not the thing itself. Left: The Treachery of Images by René Magritte (1929) makes the point that a painting of a pipe is not the thing it represents (a pipe). Right: similarly, an filled with coloured rectangles is not a barplot, unless the rectangles accurately represent some underlying data.}\label{fig:magritte}
\end{figure}

Thus, we always have to start with the data. But the raw data is rarely
something we can just ``throw'' at the computer screen. Instead, we
often first need to massage and translate it into a format that can be
understood by the graphics device. In most data visualization packages,
this is done in two steps: aggregation and scaling.

First, the data needs to be aggregated or ``wrangled'' into sets of
statistical summaries. These summaries may be very simple: for instance,
in a typical scatterplot, we assign each point the raw values of the x-
and y-variables (the summary function is then just identity,
\(f(x) = x\)). More often, however, the data is grouped and aggregated
in some way. For example, as was discussed previously, in a barplot,
data points are grouped by the levels of the x-axis variable, and each
group is then summarized by some descriptive statistic, such as count,
sum, mean, or maximum. Likewise, in a typical histogram, the x-axis
variable is divided into bins, and each bin is then summarized by the
number of cases that fall inside it.

Second, the computed statistics have to be encoded into graphical
attributes such as position, length, size, or colour \citep[see
e.g.][]{franconeri2021, wilkinson2012, pinker1990}. This is the job of
specialized functions known as scales or coordinate systems \citep[see
e.g.][]{murrell2005, murrell2007, wickham2010, wickham2016, wilkinson2012}.
For instance, in a typical scatterplot, the values of the x-axis
variable are mapped to the x-position, such that points corresponding to
small values of the x-axis variable are placed near the left-hand side
of the plot (and larger values are place near the right-hand side).
These mappings may be subject to non-linear transformations, such as log
or square-root, and may also translate discrete values, such as barplot
categories. Scales and coordinate systems are a rich subject and we have
barely scratched the surface here; interested reader is advised to see
\citet{wilkinson2012}, \citet{ziemkiewicz2009}, or \citet{hotz2020}.

@@

\begin{quote}
``This system cannot produce a meaningless graphic, however. This is a
strong claim, vulnerable to a single counter-example. It is a claim
based on the formal rules of the system, however, not on the evaluation
of specific graphics it may produce.''

``Some of the combinations of graphs and statistical methods may be
degenerate or bizarre, but there is no moral reason to restrict them.''

\citet{wilkinson2012}, The Grammar of Graphics, pp.~15 and 112.
\end{quote}

But what does stacking mean in the context of the statistics we are
trying to represent? To a data visualization novice, stacking may appear
as a purely graphical operation. If they attempt to implement a data
visualization pipeline from scratch, they may leave stacking to the very
end, after the statistical summaries have already been translated into
screen coordinates.

The view of stacking as a purely graphical operation has its appeal.
Many modern data visualization systems follow the grammar-based model of
visualization
\citep{mcnutt2022, kim2022, vanderplas2020, wickham2010, satyanarayan2014, satyanarayan2016, wilkinson2012}.
This model emphasizes compositionality, and enables the user to build
plots from independent, modular components. For example, in the popular
\texttt{ggplot2} library \citep{wickham2010}, plots are built by
combining components such as geometric objects (called \texttt{geoms})
and statistical summaries (\texttt{stats}). This grammar-based approach
offers a great deal of flexibility and expressiveness, and has gained
widespread popularity as a result.

Thus, it may be tempting to treat stacking just as another modular
component of a visualization, such that we may be able to freely stack
any geometric objects and statistics we like. Unfortunately, there is a
deep flaw in this approach. Many data visualization experts have warned
about this:

@@

The message is clear: stacking some statistics such as counts, sums, and
percentages is acceptable, but stacking others is not. As outlined by
Wills, while the sum of grouped counts produces a valid overall count,
the sum of group means does not yield a meaningful statistic. Even
averaging group means, as noted by \citet{wu2022}, does not produce a
valid summary statistic - the mean of group means differs from the grand
mean (unless by coincidence). This problem is general and applies beyond
barplots; for example, concerns about the statistical validity of
stacked density plots have also been recently raised \citep{pu2020}.

What exactly makes stacking counts and sums acceptable but stacking
other statistics such as means not? Wilke, Wills, and Wu et al.~all
argue that the aggregated value should be ``meaningful''. In the case of
sums, there is perhaps a colloquial way to put this: ``the whole is
equal to the sum of its parts''. Or, more specifically:
